---
title: "Modelling the invasive American Mink in the Pyrenees"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
# I'm going to define the function that does a simulation run up here and hide
# it, since it's not that relevant to the actual models. Of course, everyone
# will see it when they download this file, but you can just ignore this if you
# don't understand it.
run_simulation <- function(
        f,        # A function that defines the system of equations
        X,        # The initial populations
        N,        # The number of generations.
        d = 0.2   # The size of the step -- keep it small to avoid chaos.
) {
        # Define the number of simulation steps.
        n_steps <- floor(N / d)
        # Define the number of populations.
        n_pop <- length(X)
        # Define a matrix to hold all the results. Have an extra column for
        # the timestamp, and an extra row for the initial conditions.
        results <- matrix(nrow = n_steps + 1, ncol = n_pop + 1)
        # Insert the initial populations.
        results[1, ] <- c(0, X)
        # Run the simulation. Use a first-order approximation.
        for (t in 1:n_steps) {
                results[t + 1, -1] <- results[t, -1] + f(results[t, -1]) * d
                # Prevent negative populations
                results[t + 1, -1] <- ifelse(
                        results[t + 1, -1] < 0,
                        0,
                        results[t + 1, -1]
                )
                # Add the timestamp
                results[t + 1, 1] <- t * d
        }
        return(results)
}
```

## The system

In the Pyrenees, several small carnivores compete for habitat and prey. Of these, the [European mink](https://en.wikipedia.org/wiki/European_mink)
is highly threatened, due to competition with the invasive [American mink](https://en.wikipedia.org/wiki/American_mink). The population of European Mink
in the Pyrenees is one of the last remaining in Europe, and conservation of the species depends on us finding a way to mitigate
the threats in this region.

The European and American mink are both semi-aquatic [mustelids](https://en.wikipedia.org/wiki/Mustelidae) which have relatively small territories on rivers or coasts. They hunt
small rodents, birds, fish, and crustaceans. Despite the similarities that have led to their common name, they are members of distinct evolutionary
lineages, with the European mink being more closely related to weasels and ferrets than the American mink. Therefore, their similar lifestyles are an example
of [convergent evolution](https://en.wikipedia.org/wiki/Convergent_evolution).

## A simple model

To start with, let's imagine how the population of the European mink looked before the invasive species arrived. We can imagine
that there is a certain amount of mink that can be supported by the environment (the [carrying capacity](https://en.wikipedia.org/wiki/Carrying_capacity))
and that the population of mink grows when there are a lower number of individuals than the carrying capacity will support. If the population
exceeds the carrying capacity, then more individuals will die than will survive (due to limited resources) and the population will shrink.
This model is called the [logistic model](https://en.wikipedia.org/wiki/Logistic_function#In_ecology:_modeling_population_growth), and it's written like this:

$$\frac{\textrm{d}x}{\textrm{d}t} = 
        r x \left( 1 - \frac{x}{K} \right)$$

This equation tells us how the population changes over time. $x$ is the number of mink at any given time. $K$ is the carrying
capacity -- let's assume that the environment can support
1,000 individuals for now. $r$ is the per-capita [*intrinsic growth rate*](https://en.wikipedia.org/wiki/Population_dynamics#Intrinsic_rate_of_increase),
we can get the value for this from the litter size of the mink. The European mink has an average litter size of five kits ^[https://en.wikipedia.org/wiki/European_mink#Reproduction_and_development], so the per-capita
rate of increase is $2.5$ (since only females bear offspring). The term $1 - \frac{x}{K}$ decreases as $x$ gets close to $K$ -- this
means that fewer and fewer offspring survive until adulthood as we get closer to the carrying capacity -- which makes sense, since
there are fewer resources remaining.

Let's model this^[Other modellers: I'm going to use simulation models
of all these examples (even when analytic solutions exist) since I think it makes it clearer to less mathsy people what's going on.].

```{r}
logistic <- function(X) {
        # Define the parameters.
        K <- 1000
        r <- 2.5
        # Calculate the result from the model.
        res <- r * X * (1 - X / K)
        return(res)
}
```

```{r}
res <- run_simulation(
        logistic,   # Use our model.
        X = 10,     # Start with 10 individuals.
        N = 10      # Run it for 20 generations (ie. years).
)
```

```{r}
# Plot our results.
plot(
        res[, 1], res[, 2],
        type = "l",    # For a line graph
        xlab = "Number of generations",
        ylab = "Number of mink",
        main = "Growth of a mink population"
)
```

We can see that the population initially grows very quickly, before levelling out at
the carrying capacity. If we look closely, we can see that it only took four
generations to get to 1,000 individuals! The power of
[exponential growth](https://en.wikipedia.org/wiki/Exponential_growth#Biology).

**Exercise:** If the average litter size is reduced to three kits -- perhaps by some chemical pollutant --
how long will it now take the population of mink to reach the carrying capacity?

## Introducing competition

Now that we have a simple model of our population of European mink in isolation, let's see what happens when
we have an invasion of American mink. We can extend the logistic model above into a pair of equations called
the [Lotka-Volterra competition model](https://en.wikipedia.org/wiki/Competitive_Lotka%E2%80%93Volterra_equations):

$$
\\
\frac{\textrm{d}x}{\textrm{d}t} = 
        r_x x \left( 1 - \frac{x + \alpha_{xy} y}{K} \right) \\
\frac{\textrm{d}y}{\textrm{d}t} = 
        r_y y \left( 1 - \frac{y + \alpha_{yx} x}{K} \right)$$

Now we have two populations, $x$ and $y$, that interact with each other. The first equation is fairly similar to
the logistic model we saw above, but now we see the population of American mink is also contributing to reaching
the carrying capacity $K$ in the term $\frac{x + \alpha_{xy} y}{K}$ (we'll talk more about the $\alpha$ in a second).
The second equation mirrors the first, and this describes the population of the invasive mink ($y$). The equations are
broadly the same because the two mink populations behave according to the same fundamental set of mechanisms.
Here, we're assuming that the carrying capacity for the mink is the same, since they are occupying the same niche.

The two constants that are different between the equations are $r_x$ and $r_y$, and $\alpha_{xy}$ and $\alpha_{yx}$. We have two different
intrinsic growth rates since the average litter size of the American mink is slightly smaller (four kits instead of five) ^[https://en.wikipedia.org/wiki/American_mink#Reproduction_and_development]. The $\alpha$s
are called *competition coefficients*, and they capture the impact that each species has on the other. Domestic American mink are slightly
larger than the wild mink, and it has been suggested that this larger size has contributed to the
[competitive exclusion](https://en.wikipedia.org/wiki/Competitive_exclusion_principle) of the native European mink ^[https://en.wikipedia.org/wiki/American_mink#Decline_of_wild_mink]. To capture this,
let's assume that the impact on a European mink from competing with an invasive American mink is *twice as large* as competing with another
European mink -- in this case we set $\alpha_{xy} = 2$. We going to assume that for the
American mink, competition with a European mink has the same impact as competition with a member of its own species, and in this case
we set $\alpha_{yx} = 1$.

Let's write this model.

```{r}
lv_competition <- function(X) {
        # Define the parameters.
        K <- 1000
        rx <- 2.5
        ry <- 2     # Smaller litters for the invader
        a_xy <- 2   # Competitive impact on the natives
        a_yx <- 1   # Competitive impact on the invaders
        # Read the current population values
        x <- X[1]
        y <- X[2]
        # Calculate the equation results.
        res_x <- rx * x * (1 - (x + a_xy * y) / K)
        res_y <- ry * y * (1 - (y + a_yx * x) / K)
        return(c(res_x, res_y))
}
```

```{r}
res <- run_simulation(
        lv_competition,   # Use our model.
        X = c(1000, 10),  # Start with 10 invaders, and the natives at carrying capacity.
        N = 100           # Run it for 100 generations (ie. years).
)
```

```{r}
# Plot our results.
plot(
        res[, 1], res[, 2],  # Natives
        type = "l",    # For a line graph
        col = "red",
        xlab = "Number of generations",
        ylab = "Number of mink",
        main = "Competition between European and American mink",
        ylim = c(0, 1000)    # Plot the full y-axis
)
lines(res[, 1], res[, 3], col = "blue")  # Invaders
legend(
        "topright",
        legend = c("European mink", "American mink"),
        fill = c("red", "blue")
)
```

Wow! The American mink starts off slowly, but around generation 45 its growth explodes, and
it drives the European mink to extinction within only a few years. There's a few things we can think about
here.

Firstly -- except for the reproductive rates -- we've guesstimated the values that go into this model ($K$ and the $\alpha$s).
If we had data on field surveys of these species, or laboratory experiments on how they compete, we could derive values for these constants
using statistical methods. For the predictions we make from this model to be meaningful, we need this data!
However, this doesn't mean our toy model is useless. We can use it to investigate the impacts that changing certain values has on
the shape of the curves that we draw. We can see some of these questions in the exercises below.

You might have noticed that for the early stages of the invasion, the population of the invasive is very low. Let's see how low:

```{r}
summary(res[
        res[, 1] < 20,   # Look at population within the first 20 years
        3                # For the invasive
])
```

Looks like for 20 years the population doesn't even reach 16 individuals! At this stage there are just too many native mink taking up space
and resources for the invasives to grow rapidly. At this point, nearly half of all the offspring of the invader die before they mature.
You might have noticed here that our model assumes that there are no fluctuations in the population -- but we know this is not realistic.
If we imagine that there is a severe drought in year 10, it is very likely that the small invasive population will be eliminated, whereas
the native population is large enough that it can survive (even with many deaths). This is also informative for planning control measures --
if we can recapture 15 wild American mink within 20 years, we can remove the threat of extinction altogether!

Finally, this model allows us to incorporate competition, but it is not explicit about the *mechanisms* by which this competition operates.
Do the American mink have a negative impact on the European mink because they competing for prey? Or for burrows? Or for territory? Perhaps
there is hybridisation? Or predation? Each of these mechanisms might have very different responses to control measures, so it can be very
important that our models have some more detail about how these mechanisms interact. In the final section we'll look at this some more.

**Excercise:** Try adjusting the value we've given to the competition coefficient for the native European mink. Is there a
value of the competition coefficient that allows these species to coexist? How does adjusting the coefficient change the time to
extinction?

**Exercise:** Imagine that through a capture-neuter-release program we are able to reduce the average litter size of the invasive American
mink from four to three kits per litter. Does this prevent extinction of the European mink?

### Extending the Lotka-Volterra model

As well as the two mink, other species with similar ranges include the [red fox](https://en.wikipedia.org/wiki/Red_fox) and
[European badger](https://en.wikipedia.org/wiki/European_badger). Despite their similar size and some overlap in food sources, these species
have importantly different [ecological niches](https://en.wikipedia.org/wiki/Ecological_niche) to the mink, with both being omnivorous and terrestrial.
It is possible to include these species in the model, and also to adjust the competition coefficients to take account of the difference in diet
and behaviour which would mean that they interact less with the species of interest. However, the model with these additions functions similarly,
at the cost of being more complicated, so we won't bother going into it here. If you're interested in how we would go about doing this, ask me.

## Propagule pressure and conservation

In this section, we will pivot a little to look at a modern model of invasion, taken from
[Catford et al. 2018](http://www.nature.com/articles/s41467-018-04491-3).

The reason that the American mink was introduced to Europe is in order to [farm it for fur](https://en.wikipedia.org/wiki/Fur_farming), and 
and escaped mink from fur farms can bolster self-sustaining wild populations. This effect is called [propagule pressure](https://en.wikipedia.org/wiki/Propagule_pressure)
in invasion science. We will incorporate the propagule pressure in this model as a new parameter $h$ for both populations -- an equivalent
for the endangered native could be a breed and release conservation project.

In this model, we will make the idea of a carrying capacity more concrete by imagining that the area of interest is divided into
territories that are occupied by a single member of one of the species. In the case of the mink, these territories are usually 1-6km of
riverbank ^[https://en.wikipedia.org/wiki/European_mink#Territorial_and_denning_behaviours],
but can also include ponds and coastal areas. The carrying capacity is the number of territories available, and instead of modelling the
number of individuals, we will model the number of territories occupied by each species (assuming that if an individual doesn't have a
territory of its own, it will die).

Additionally, we will add a new
term that represents per-capita mortality of the mink. Both species have a lifespan of approximately 10 years (although the 
invasive may have a longer lifespan), so the per-capita mortality
rate is $m_x = m_y = \frac{1}{10} = 0.1$ (we assume that the only cause of death of an individual which occupies some habitat is
[senescence](https://en.wikipedia.org/wiki/Senescence))

Finally, we will think about competition more concretely. There is some evidence that species around the world make
*universal trade-offs* (see [Tilman 2011](https://www.journals.uchicago.edu/doi/10.1086/661245)). That is, every species is subject to
fundamental biological constraints that limit how many resources they can invest in different axes of competition. In this model, we will
assume that there is a tradeoff between fecundity (litter size) and ability to compete for habitat. In this
case, we could think of this in terms of resource use between the two species. We know that the European mink produces more offspring and
that the American mink is slightly larger, representing a tradeoff in the allocation of
[biomass](https://en.wikipedia.org/wiki/Biomass_(ecology)) acquired from food. Let's assume that the larger size of the invasive allows it
to displace European mink from habitat patches they occupy.

Let's build our model.

$$
\\
\frac{\textrm{d}x}{\textrm{d}t} = 
        \left( r_x x + h_x \right) \left( 1 - \frac{x + y}{K} \right) -
        \left( m_x + r_y \frac{y}{K} + h_y \right) x \\
\frac{\textrm{d}y}{\textrm{d}t} = 
        \left( r_y y + h_y \right) \left( 1 - \frac{y}{K} \right) -
        m_y y$$

Now, these equations are different for the two species (due to differences we outlined above). Each equation has two parts, the first
representing colonisation of new territories, and the second part representing loss of territory due to death or displacement. In the case
of the native European mink ($x$), the proportion of patches avaiable to it for colonisation is

```{r}
catford <- function(
        X,
        hx = 0,
        hy = 0,    # Propagule pressure -- 0 for now.
        mx = 0.1,
        my = mx   # Mortality rates -- equal for now.
) {
        # Define the parameters.
        K <- 100    # Number of habitat patches
        rx <- 2.5
        ry <- 2     # Smaller litters for the invader
        # Read the current population values
        x <- X[1]
        y <- X[2]
        # Calculate the equation results.
        res_x <- (rx * x + hx) * (1 - (x + y) / K) - (mx + ry * y / K + hy) * x
        res_y <- (ry * y + hy) * (1 - y / K) - my * y
        return(c(res_x, res_y))
}
```

```{r}
res <- run_simulation(
        catford,          # Use our model.
        X = c(100, 10),   # Start with 10 invaders, and the natives at carrying capacity.
        N = 20            # Run it for 20 generations (ie. years).
)
```

```{r}
# Plot our results.
plot(
        res[, 1], res[, 2],  # Natives
        type = "l",    # For a line graph
        col = "red",
        xlab = "Number of generations",
        ylab = "Number of mink",
        main = "Competition between European and American mink",
        ylim = c(0, 100)    # Plot the full y-axis
)
lines(res[, 1], res[, 3], col = "blue")  # Invaders
legend(
        "topright",
        legend = c("European mink", "American mink"),
        fill = c("red", "blue")
)
```

Wow, so in this model, we get extinction of the native almost immediately (after only a couple of years). Now, let's think about our
breed-and-release program -- how many European mink would we need to breed and release every year to have a viable population of the native?

```{r}
# Let's run a simulation for values of hx ranging from 1 to 1000, and see
# how many European mink-occupied habitats we have after 50 years.
hx <- 1:1000
res <- rep(NA, length(hx))    # create a vector to hold the results
for (i in seq_along(hx)) {
        sim_res <- run_simulation(
                # Use our model with the new value of h
                function(X) catford(X, hx = hx[i]),
                X = c(100, 10),   # Start with 10 invaders, and the natives at carrying capacity.
                N = 50            # Run it for 50 generations (ie. years).
        )
        # Store the last value of the native population.
        res[i] <- floor(sim_res[nrow(sim_res), 2])
}
```

```{r}
# Let's plot the results.
plot(
        hx, res,
        xlab = "Size of annual release",
        ylab = "Stable native population size",
        main = "Size of breeding program necessary to maintain populations of the native after 50 years",
)
```

Looking at this, we can see that even when we breed and release hundreds of native mink a year, we are unable to maintain a number of
occupied patches more than 10. If there are also American mink escaping from fur farms (i.e. if $h_y > 0$), this will be even less
effective.

Another way that we can improve the chances of the native mink is to increase the mortality rate of the invader. Since the only habitat
patches that the native species is able to colonise are those which are empty (i.e. those vacated by death), increasing the rate of death
increases the number of patches available to the native. This could be achieved via a culling program. We will look at this in the exercise.

There are some problems with this model. Firstly, it's somewhat unrealistic to assume that the American mink will *always* be able to displace
the European mink if they compete for habitat. The invasive may be larger *on average*, but there will be cases where a larger-than-average
European mink is able to displace (or resist displacement by) an American mink. We could incorporate this into the model by allowing the
natives to colonise some territories occupied by the invasive, and also by preventing the invasive from colonising *all* of the territories
held by the native. Secondly, we know that human-caused disturbance will be impacting these populations too -- we should really account for
this. Also, it seems presumptuous to assume that the mortality rate of the mink is not part of the tradeoff that they make. I (Jonathan)
will be incorporating all of these factors (and more) into this model as part of my PhD!

**Exercise:** What level of mortality for the invader produces a stable population of the native (hint: use the same method as we used above
for the breed-and-release program)?

**Exercise:** What if we have differing mortality as well as breed-and-release? Can you come up with a method to do a cost-benefit
analysis between culling and breed-and-release?