---
title: "Modelling the invasive American Mink in the Pyrenees"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The model

This model is taken from [Catford et al. 2018](http://www.nature.com/articles/s41467-018-04491-3).

$$\frac{\textrm{d}p_i}{\textrm{d}t} = 
        \left( c_i p_i + h_i \right) \left( 1 - \sum_{j = 1}^{i} p_j \right) -
        \left( m_i + \sum_{j = 1}^{i-1} c_j p_j + h_j \right) p_i$$



## GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Including Code

You can include R code in the document as follows:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
