---
title: "Modelling the invasive American Mink in the Pyrenees"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The system

In the Pyrenees, several small carnivores compete for habitat and prey. Of these, the [European Mink](https://en.wikipedia.org/wiki/European_mink)
is highly threatened, due to competition with the invasive [American Mink](https://en.wikipedia.org/wiki/American_mink). The population of European Mink
in the Pyrenees is one of the last remaining in Europe, and conservation of the species depends on us finding a way to mitigate
the threats in this region.

The European and American mink are both semi-aquatic [mustelids](https://en.wikipedia.org/wiki/Mustelidae) which have relatively small territories on rivers or coasts. They hunt
small rodents, birds, fish, and crustaceans. Despite the similarities that have led to their common name, they are members of distinct evolutionary
lineages, with the European mink being more closely related to weasels and ferrets than the American mink. Therefore, their similar lifestyles are an example
of [convergent evolution](https://en.wikipedia.org/wiki/Convergent_evolution).

As well as the two mink, other species with similar ranges include the [red fox](https://en.wikipedia.org/wiki/Red_fox) and
[European badger](https://en.wikipedia.org/wiki/European_badger). Despite their similar size and some overlap in food sources, these species
have importantly different [ecological niches](https://en.wikipedia.org/wiki/Ecological_niche) to the mink, with both being omnivorous and terrestrial. We will include these species in the model to make it more realistic, but taking
account of the difference in diet and behaviour which should mean that they interact less with the species of interest.

## The model

This model is taken from [Catford et al. 2018](http://www.nature.com/articles/s41467-018-04491-3).

$$\frac{\textrm{d}p_i}{\textrm{d}t} = 
        \left( c_i p_i + h_i \right) \left( 1 - \sum_{j = 1}^{i} p_j \right) -
        \left( m_i + \sum_{j = 1}^{i-1} c_j p_j + h_j \right) p_i$$



## GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Including Code

You can include R code in the document as follows:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
