---
title: "Modelling the invasive American Mink in the Pyrenees"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The system

In the Pyrenees, several small carnivores compete for habitat and prey. Of these, the [European mink](https://en.wikipedia.org/wiki/European_mink)
is highly threatened, due to competition with the invasive [American mink](https://en.wikipedia.org/wiki/American_mink). The population of European Mink
in the Pyrenees is one of the last remaining in Europe, and conservation of the species depends on us finding a way to mitigate
the threats in this region.

The European and American mink are both semi-aquatic [mustelids](https://en.wikipedia.org/wiki/Mustelidae) which have relatively small territories on rivers or coasts. They hunt
small rodents, birds, fish, and crustaceans. Despite the similarities that have led to their common name, they are members of distinct evolutionary
lineages, with the European mink being more closely related to weasels and ferrets than the American mink. Therefore, their similar lifestyles are an example
of [convergent evolution](https://en.wikipedia.org/wiki/Convergent_evolution).

## [Lotka-Volterra competition model](https://en.wikipedia.org/wiki/Competitive_Lotka%E2%80%93Volterra_equations)

$$\frac{\textrm{d}x_i}{\textrm{d}t} = 
        r x_i \left( 1 - \frac{x_i + \sum_{j \neq i} \alpha_{ij} x_j }{K_i} \right)$$



## Escapes from farms

The reason that the American mink was introduced to Europe is in order to [farm it for fur](https://en.wikipedia.org/wiki/Fur_farming), and 
and escaped mink from fur farms can bolster 

*This model is taken from [Catford et al. 2018](http://www.nature.com/articles/s41467-018-04491-3)*.

$$\frac{\textrm{d}p_i}{\textrm{d}t} = 
        \left( c_i p_i + h_i \right) \left( 1 - \sum_{j = 1}^{i} p_j \right) -
        \left( m_i + \sum_{j = 1}^{i-1} c_j p_j + h_j \right) p_i$$

In this model, we will imagine that the area of interest is divided into territories that are occupied by a single member of one of the species.
In the case of the mink, these territories are usually 1-6km of riverbank, but can also include ponds and coastal areas. We will model the proportion
of the region that is occupied by each of these species. For example, if European mink occupies 90% of the available habitat, then the American
mink will occupy 10%.

The female American mink has four kits on average per year, and the European mink has a slightly larger litter of five kits. There is
evidence that species around the world make *universal trade-offs* (see [Tilman 2011](https://www.journals.uchicago.edu/doi/10.1086/661245)),
that is, for every 

## Extending the model

As well as the two mink, other species with similar ranges include the [red fox](https://en.wikipedia.org/wiki/Red_fox) and
[European badger](https://en.wikipedia.org/wiki/European_badger). Despite their similar size and some overlap in food sources, these species
have importantly different [ecological niches](https://en.wikipedia.org/wiki/Ecological_niche) to the mink, with both being omnivorous and terrestrial.
It is possible to include these species in the model, and also to adjust it to take
account of the difference in diet and behaviour which would mean that they interact less with the species of interest (**TODO:** incorporate this?).


$$\frac{\textrm{d}p_i}{\textrm{d}t} = 
        \left( c_i p_i + h_i \right) \left( 1 - \sum_{j = 1}^{i} p_j \right) -
        \left( m_i + \sum_{j = 1}^{i-1} c_j p_j + h_j \right) p_i$$

## GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Including Code

You can include R code in the document as follows:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
